---
title: "Biostat 203B Homework 1 Solutions"
author: Olivia Golston
subtitle: Due Jan 22 @ 11:59PM
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Display machine information for reproducibility:
```{r, eval=F}
sessionInfo()
```

## Q1. Git/GitHub

**No handwritten homework reports are accepted for this course.** We work with Git and GitHub. Efficient and abundant use of Git, e.g., frequent and well-documented commits, is an important criterion for grading your homework.

1. Apply for the [Student Developer Pack](https://education.github.com/pack) at GitHub using your UCLA email.

**Solution**: Done. 

2. Create a **private** repository `biostat-203b-2021-winter` and add `Hua-Zhou`, `Chris-German` and `ElvisCuiHan` as your collaborators with write permission.

**Solution**: Done

3. Top directories of the repository should be `hw1`, `hw2`, ... Maintain two branches `master` and `develop`. The `develop` branch will be your main playground, the place where you develop solution (code) to homework problems and write up report. The `master` branch will be your presentation area. Submit your homework files (R markdown file `Rmd`, `html` file converted from R markdown, all code and data sets to reproduce results) in `master` branch.

**Solution**: Done

4. After each homework due date, teaching assistant and instructor will check out your master branch for grading. Tag each of your homework submissions with tag names `hw1`, `hw2`, ... Tagging time will be used as your submission time. That means if you tag your `hw1` submission after deadline, penalty points will be deducted for late submission.

**Solution**: Done


5. After this course, you can make this repository public and use it to demonstrate your skill sets on job market.

**Solution**: Will do.


## Q2. Linux Shell Commands

1. This exercise (and later in this course) uses the [MIMIC-IV data](https://mimic-iv.mit.edu), a freely accessible critical care database developed by the MIT Lab for Computational Physiology. Follow the instructions at <https://mimic-iv.mit.edu/docs/access/> to (1) complete the CITI `Data or Specimens Only Research` course and (2) obtain the PhysioNet credential for using the MIMIC-IV data. Display the verification links to your completion report and completion certificate here. (Hint: The CITI training takes a couple hours and the PhysioNet credentialing takes a couple days; do not leave it to the last minute.)


**Solution**: 

For the basic course: 
Here is my [course completion report](http://citiprogram.org/verify/?k1d1732f0-9f35-422d-9a58-d821623e5cb5-40343731).
Here is my [course completion certificate](http://citiprogram.org/verify/?w2fa5babf-42a9-4f6d-8012-fd080b812c7d-40343731).

Due to an issue regarding some of my coursework being too old, I was instructed to complete the refresher course in order to get PhysioNet credentials:
Here is my [course completion report](http://citiprogram.org/verify/?k042b311f-9e6e-4456-970f-6b68e979fd7b-40380846).
Here is my [course completion certificate](http://citiprogram.org/verify/?w90e40897-5d20-4719-bd72-625d3b6a9492-40380846).


2. The `/usr/203b-data/mimic-iv/` folder on teaching server contains data sets from MIMIC-IV. Refer to <https://mimic-iv.mit.edu/docs/datasets/> for details of data files.  

```{bash}
  ls -l /usr/203b-data/mimic-iv
```
Please, do **not** put these data files into Git; they are big. Do **not** copy them into your directory. Do **not** decompress the gz data files. These create unnecessary big files on storage and are not big data friendly practices. Just read from the data folder `/usr/203b-data/mimic-iv` directly in following exercises. 

    Use Bash commands to answer following questions.

3. Display the contents in the folders `core`, `hosp`, `icu`. What are the functionalities of the bash commands `zcat`, `zless`, `zmore`, and `zgrep`? 


**Solution**:  

```{bash, eval=F}
  ls -l /usr/203b-data/mimic-iv/core/
```
    
```{bash eval=F}
  ls -l /usr/203b-data/mimic-iv/hosp/
```
    
```{bash eval=F}
  ls -l /usr/203b-data/mimic-iv/icu/
```

`zcat`, `zless`, `zmore`, and `zgrep` function like `cat`, `less`, `more`, and `grep`, but work for compressed data files.

```{bash, eval=F}
   zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | head
```

`zcat` prints the contents of a compressed file


```{bash, eval=F}
   zless /usr/203b-data/mimic-iv/core/admissions.csv.gz
```

`zless` allows you to browse a compressed file using the functionality of `less`


```{bash, eval=F}
   zmore /usr/203b-data/mimic-iv/core/admissions.csv.gz
```

`zmore` allows you to browse a compressed file using the functionality of `more`

```{bash, eval=F}
   zgrep 'PHYSICIAN' /usr/203b-data/mimic-iv/core/admissions.csv.gz | head
```

`zgrep` allows you to search for certain characters within a compressed file



4. What's the output of following bash script?
    ```{bash, eval=F}
    for datafile in /usr/203b-data/mimic-iv/core/*.gz
      do
        ls -l $datafile
      done
    ```
Display the number of lines in each data file using a similar loop.

**Solution**: the given loop outputs the files found within the "core" directory, along with the permissions and other information. The three files in the core directory are admissions.csv, patients.csv, and transfers.csv. They all have -rw-r--r-- permissions.  

    ```{bash eval=F}
    for datafile in /usr/203b-data/mimic-iv/core/*.gz
      do
        zcat $datafile | wc -l
      done
    ```
**Solution** There are 524521, 383221, and 2192964 lines respectively. 

4. Display the first few lines of `admissions.csv.gz`. How many rows are in this data file? How many unique patients (identified by `subject_id`) are in this data file? What are the possible values taken by each of the variable `admission_type`, `admission_location`, `insurance`, `language`, `marital_status`, and `ethnicity`? Also report the count for each unique value of these variables. (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `uniq`, `wc`, and so on.)

```{bash eval=F}
  zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | head
```
    
Columns in order : 
1) subject_id
2) hadm_id
3) admittime
4) dischtime
5) deathtime
6) admission_type
7) admission_location
8) discharge_location
9) insurance
10) language
11) marital_status
12) ethnicity
13) edregtime
14) edouttime
15) hospital_expire_flag    
    
```{bash eval=F}
  zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | wc -l
```
There are 524521 lines.


```{bash eval=F}
  zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, ' { print $1 }' | 
  sort -u | wc -l
```
```{bash eval=F}
  zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, ' { print $1 }' | 
  sort | head
```


### Admission Type
```{bash eval=F}
   zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, '{if (FNR > 1) print $6 }' | 
   sort | uniq -c 
  
```

### Admission Location
```{bash eval=F}
  zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, '{if (FNR > 1) print $7 }' | 
  sort | uniq -c 
```

### Insurance
```{bash eval=F}
  zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, '{if (FNR > 1) print $9 }' | 
  sort | uniq -c 
```
### Language
```{bash eval=F}
  zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, '{if (FNR > 1) print $10 }' | 
  sort | uniq -c 
```


### Marital Status
```{bash eval=F}
  zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, '{if (FNR > 1) print $11 }' | 
  sort | uniq -c 
```


### Ethnicity
```{bash eval=F}
  zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, '{if (FNR > 1) print $12 }' | 
  sort | uniq -c 
```

## Q3. Who's popular in Price and Prejudice

1. You and your friend just have finished reading *Pride and Prejudice* by Jane Austen. Among the four main characters in the book, Elizabeth, Jane, Lydia, and Darcy, your friend thinks that Darcy was the most mentioned. You, however, are certain it was Elizabeth. Obtain the full text of the novel from <http://www.gutenberg.org/cache/epub/42671/pg42671.txt> and save to your local folder. 
    ```{bash, eval=FALSE}
    curl http://www.gutenberg.org/cache/epub/42671/pg42671.txt > pride_and_prejudice.txt
    ```
Do **not** put this text file `pride_and_prejudice.txt` in Git. Using a `for` loop, how would you tabulate the number of times each of the four characters is mentioned?

```{bash eval=F}
  for name in "Elizabeth" "Jane" "Lydia" "Darcy"
    do
      echo $name:
      grep -io $name pride_and_prejudice.txt | wc -l
    done

```



2. What's the difference between the following two commands?
    ```{bash eval=FALSE}
    echo 'hello, world' > test1.txt
    ```
    and
    ```{bash eval=FALSE}
    echo 'hello, world' >> test2.txt
    ```


**Solution**: The command `echo 'hello, world' > test1.txt ` writes 'hello, world' to the test1 text file. If there is already text in the file, it will be overwritten. Thus, the result of this command - no matter how many times it is run - will be a file with 'hello, world' written one time.  
The command `echo 'hello, world' >> test2.txt` appends 'hello, world' to the test2 text file. If there is already text in the file, it will **not** be overwritten. Thus, the result of running this command 4 times would be a file with 'hello, world' written 4 times. 

3. Using your favorite text editor (e.g., `vi`), type the following and save the file as `middle.sh`:
    ```{bash eval=FALSE}
    #!/bin/sh
    # Select lines from the middle of a file.
    # Usage: bash middle.sh filename end_line num_lines
    head -n "$2" "$1" | tail -n "$3"
    ```
Using `chmod` make the file executable by the owner, and run 
    ```{bash eval=FALSE}
    ./middle.sh pride_and_prejudice.txt 20 5
    ```
Explain the output. Explain the meaning of `"$1"`, `"$2"`, and `"$3"` in this shell script. Why do we need the first line of the shell script?

**Solution**: The output is the Author and Editor of Pride and Prejudice. The first line is included to specify what shell the program needs to be executed with (sh). This might be especially important if using the script on a different system, which may be using a different shell by default. 

`"$1"`, `"$2"`, and `"$3"` refer to the arguments for the script. The resulting program takes the first `$2` (in this case, 20) lines of file `"$1"` (`pride_and_prejudice.txt`). These lines are piped to the next command, which takes the last `$3` (5) lines.  For these particular arguments, the result is the 15th through 20th lines of the Pride and Prejudice text. 

### Q4. More fun with Linux

Try these commands in Bash and interpret the results: `cal`, `cal 2021`, `cal 9 1752` (anything unusual?), `date`, `hostname`, `arch`, `uname -a`, `uptime`, `who am i`, `who`, `w`, `id`, `last | head`, `echo {con,pre}{sent,fer}{s,ed}`, `time sleep 5`, `history | tail`.


**Solution**: 
```{bash}
cal
```
`cal` displays the calendar for the current month. 


```{bash}
cal 2021
```
`cal 2021` calendar for the year 2021. Substituting another year would show the calendar for the specified year.

```{bash}
cal 9 1752
```
`cal 9 1752` displays a calendar for the requested month in the requested year. Strangely, the month of September in 1752 had 11 days missing! This was done during the switch to the Gregorian calendar.

```{bash}
date
```
`date` gives the current day of the week, month, day, and time in Coordinated Universal Time.

```{bash}
hostname
```
`hostname` gives the name of the server. In this case, that is ucla-biostat-203b

```{bash}
arch
```
`arch` prints out the computer architecture. For me, this is x86_64.

```{bash}
uname -a
```
`uname -a` prints out detail about the current user, server, and timing. For example, it will print the operating system, server/network name, computer architecture, and date.


```{bash}
uptime
```

`uptime` gives some statistics about the system. Specifically, the current time (in UTC), followed by the amount of the time the system has been up, the number of logged in users, and system load averages.


```{bash}
who am i
```

`who am i` ???


```{bash}
who
```

`who`


```{bash}
w
```
`w`

```{bash}
id
```

`id`


```{bash}
last | head
```
`last | head` gives the most recent 10 entries (`head`) from a list of logins into the server (given by `last`). This list gives usernames, date, log-in and log-out times, and other information. 

```{bash}
echo {con,pre}{sent,fer}{s,ed}
```
`echo {con,pre}{sent,fer}{s,ed}` prints all 8 combinations the given word parts. There are three "positions", each with two possible entries, for a total of `2 * 2 * 2 = 8` total words to display.  

```{bash echo=F}
time sleep 5
```
`time sleep 5` tells the system to sleep for 5 seconds (`sleep 5`), and then prints the time that it took to do that command (`time`).


```{bash echo=F}
history | tail
```
`history | tail` gives the most recent 10 commands that I have executed. As the most recently typed command, `history | tail` is at the bottom of this list.    
